Continous time Markov chain models; \\ extension, simulation and visualisation

### 2014-06-04 PhDThesis.tex
finished working on M-H description. Now on to the examples

%%%%%%%%%%%%%%%%%%
%---TITLE PAGE---%
%%%%%%%%%%%%%%%%%%
\title{Continous time Markov chain models in phylogenetics and phylogeography; model extension, simulation and visualisation}

@@

\subsection{Computational burden of the likelihood computation}
% https://class.coursera.org/algo-004/lecture/11

@@

As a simple illustration of the concept of Bayesian analysis let us consider a toy example where two sequences are compared and $x=20$ differences are found out of $N=1000$ sites compared.
Suppose that the parameter $\theta$ we want to infer is the proportion of sites which differ between the two sequences.
Assuming that the sites are independent we can use a binomial model and we put a beta prior with parameters $\alpha=2$, $\beta=100$ on the unknown proportion $\theta$.

%---LIKELIHOOD---%
\begin{figure}[h]
\begin{center}
\includegraphics[scale=0.5]{bayes} 
\end{center}
\caption{{ \footnotesize {\bf Beta prior and posterior densities plotted with a scaled likelihood resulting from a binomial model.} 
}}
\label{fig:bayes1}
\end{figure}

Figure \ref{fig:bayes1} shows the resulting posterior distribution along with prior and scaled likelihood.

@@

Metropolis %-Hasting 
algorithm, as proposed by \citet{Metropolis1953}, is an MCMC algorithms for sampling from multi-dimensional distributions.

% Common choice is to sample from a continuous uniform distribution centered around the current state of the chain: 
% $$\theta^{*}\sim U\left[\theta[t]-\omega,\theta[t]+\omega\right],$$
% where $\omega$ is the arbitrary window size.

\begin{algorithm}
\begin{center}
\begin{algorithmic}[1]
\footnotesize{
%
\State Pick a starting value of the chain $\theta \left[ t \right]$ for $t=0$.
%
\For{int $\left(t=0; \; t<\text{chain length}; \; t++\right)$}
%
\State Generate a candidate state from the proposal distribution (kernel):
$$\theta^{*} \sim q(\theta[t])$$
%
\State Calculate the acceptance ratio:
$$r=\frac{prior\left(\theta^{*}\right)}{prior\left(\theta\left[t\right]\right)}\cdot\frac{Likelihood\left(\mathbf{X}|\theta^{*}\right)}{Likelihood\left(\mathbf{X}|\theta\left[t\right]\right)}.$$
%
\State Sample $u\sim U[0,1]$.
%
\If{$\left( u < r \right)$}
%
\State Set $\theta[t+1]=\theta^*.$
%
\Else 
%
\State Set $\theta[t+1]=\theta[t].$
%
\EndIf
%
\EndFor
}
\end{algorithmic}
\end{center}
\caption{{ \footnotesize {\bf AAA} 
}}
\end{algorithm}

@@

%plots of gibbs sampler path etc

>>>>

%%%%%%%%%%%%%%%%%%
%---TITLE PAGE---%
%%%%%%%%%%%%%%%%%%
\title{Continous time Markov chain models; \\
% in phylogenetics and phylogeography; model 
extension, simulation and visualisation}

@@

\subsection{Computational burden of the likelihood computation}

@@

As a simple illustration of the concept of Bayesian analysis let us consider a toy example where two sequences are compared and $x=10$ differences are found out of $N=100$ sites compared.
Suppose that the parameter $\theta$ we want to infer is the proportion of sites which differ between the two sequences.
Assuming that the sites are independent we can use a binomial model and we put a beta prior with parameters $\alpha=2$, $\beta=100$ on the unknown proportion $\theta$.

%---LIKELIHOOD---%
\begin{figure}[H]
\begin{center}
\includegraphics[scale=0.5]{bayes} 
\end{center}
\caption{{ \footnotesize {\bf Beta prior and posterior densities plotted with a scaled likelihood resulting from a binomial model.} 
}}
\label{fig:bayes1}
\end{figure}

Figure \ref{fig:bayes1} shows the resulting posterior distribution along with prior and scaled likelihood.
In this case we used a fact that the beta distribution is a so called \emph{conjugate prior} for the binomial model, resulting in the posterior which is also beta-distributed and can be calculated analytically.
%TODO: proof in appendix?

@@

%TODO: word for word, description
Metropolis-Hasting 
algorithm, as proposed by \citet{Metropolis1953}, is an MCMC algorithms for sampling from multi-dimensional distributions.
It proceeds by constructing a Markov chain with its states being the values of the parameter of interest $\theta$ and whose equilibrium distribution is the posterior $P\left(\theta|\mathbf{X}\right)$ as defined in Equation \ref{eq:bayesian1}.
After choosing an arbitrary starting state for the chain new candidate states $\theta^{*}$ are generated from the proposal density $q(\theta^{*} | \theta[t])$ also called the kernel.

Common choice is to sample from a continuous uniform distribution centered around the current state of the chain, the so called jumping kernel:

$$\theta^{*}\sim U\left[\theta[t]-\omega,\theta[t]+\omega\right],$$

\noindent
where $\omega$ is the arbitrary window size.
The chain then moves to the proposed value with probability defined by the acceptance ratio (see Equation \ref{eq:metropolis1}).
If the proposal is rejected the chains remains at it's current state and the routine iterates between accept/reject steps a specified number of times.
Both acceptance and rejection are counted as iterations od the algorithm, in fact the rejections ensure that the probabilistic properties of the target distribution are well represented in the generated sample.  
Listing in Algorithm \ref{alg:metropolisHastings} presents the details of the Metropolis-Hastings.

\begin{algorithm}
\begin{center}
\begin{algorithmic}[1]
\footnotesize{
%
\State Pick a starting value of the chain $\theta \left[ t \right]$ for $t=0$.
%
\For{int $\left(t=0; \; t<\text{chain length}; \; t++\right)$}
%
\State Generate a candidate state from the proposal distribution (kernel):
$$\theta^{*} \sim q(\theta[t])$$
%
\State Calculate the acceptance ratio:
$$r=\frac{prior\left(\theta^{*}\right)}{prior\left(\theta\left[t\right]\right)}\cdot\frac{Likelihood\left(\mathbf{X}|\theta^{*}\right)}{Likelihood\left(\mathbf{X}|\theta\left[t\right]\right)}\cdot\frac{q\left(\theta\left[t\right]|\theta^{*}\right)}{q\left(\theta^{*}|\theta\left[t\right]\right)}.$$
%
\State Sample $u\sim U[0,1]$.
%
\If{$\left( u < r \right)$}
%
\State Set $\theta[t+1]=\theta^*.$
%
\Else 
%
\State Set $\theta[t+1]=\theta[t].$
%
\EndIf
%
\EndFor
}
\end{algorithmic}
\end{center}
\caption{{ \footnotesize {\bf Metropolis-Hastings algorithm} 
}}
\label{alg:metropolisHastings}
\end{algorithm}

We need to note several important features of the algorithm.
The most important part of the algorithm is the proposal density. 
Given the current state it proposes the next state independent of the past states.
We can recall that this is the Markov property, in fact the output of the algorithm generates a Markov chain (hence the name).
Metropolis-Hastings algorithm allows for the use of asymetrical proposal densities such that $q\left(\theta\left[t\right]|\theta^{*}\right)\neq q\left(\theta^{*}|\theta\left[t\right]\right)$.
The acceptance probability is simply the ratio of the target posterior evaluated at the proposal and at the current state of the chain, times the ratio of the proposal: 
% prior ratio times the likelihood ratio times the proposal ratio:

\begin{equation}
\frac{P\left(\theta^{*}|\mathbf{X}\right)\cdot q\left(\theta\left[t\right]|\theta^{*}\right)}{P\left(\theta\left[t\right]|\mathbf{X}\right)\cdot q\left(\theta\left[t\right]|\theta^{*}\right)}=\frac{P(\theta^{*})\cdot P\left(\mathbf{X}|\theta^{*}\right)\cdot q\left(\theta\left[t\right]|\theta^{*}\right)}{P(\theta\left[t\right])\cdot P\left(\mathbf{X}|\theta\left[t\right]\right)\cdot q\left(\theta^{*}|\theta\left[t\right]\right)}.
\label{eq:metropolis1}
\end{equation}

Note that the integral in (\ref{eq:bayesian1}) which is notoriously hard to compute cancels out in (\ref{eq:metropolis1}).
Thus to estimate the posterior $P\left(\mathbf{\theta}|\mathbf{X}\right)$ one only needs to run the algorithm for a sufficiently long time to let the chain reach it's equilibrium, which is precisely the target posterior.

@@

%TODO: explain probabilities

%---METROPOLIS PATH---%
\begin{figure}[H]
\begin{center}
\includegraphics[scale=0.5]{metropolis} 
\end{center}
\caption{{ \footnotesize {\bf AAA.} 
}}
\label{fig:metropolis}
\end{figure}
